\documentclass[nohyper]{tufte-handout}
% The hyperref is loaded in the preamble with additional arguments to
% avoid some xelatex warnings.

\title[Asymptotic statistics]%
{Asymptotic Statistics \\
  Core Econometrics Notes, part 3}
\input{common/preamble}

\begin{document}
\maketitle

\bigskip\noindent%
Copyright © 2013, the authors of \textit{Core Econometrics Notes}.
A complete author list is included in Appendix A of this document.

Permission is granted to copy, distribute and/or modify this document
under the terms of the GNU Free Documentation License, Version 1.3 or
any later version published by the Free Software Foundation; with no
Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts.  A
copy of the license is included in Appendix B of this document and is
also available online at \url{http://www.gnu.org/copyleft/fdl.html}.

This text was produced as part of the Econometrics Free Library
Project.  The project's goal is to produce high-quality free and
open access econometrics textbooks and reference material.  More
information about this project is available at its homepage,
\homepage, including information on how
to participate.  This text is typeset in \LaTeX\ and there is a link
to the source of the document at the project's homepage as well.

Please cite this document as
\begin{itemize}
\item[] \bibentry{eflp-core}
\item[] \verbatiminput{CITATION.bib}
\end{itemize}
The \BibTeX\ record is provided for your convenience.
If the year and version are not specified, please see the files
\texttt{README.md} or \texttt{VERSION.tex} for directions on how to
find them.

If you find errors in this text, please email the project mailing
list, \maillist, or log the error in the project issue tracker,
\bugtrack.  It would be helpful if you could refer to the version
number listed above as well.  Thanks!

\addcontentsline{toc}{part}{Table of Contents}
\tableofcontents

\include{asymptotics/law_of_large_numbers}
\include{asymptotics/central_limit_theorem}
\include{asymptotics/point_estimation}
\include{asymptotics/linear_regression}
\include{asymptotics/inference}
\include{asymptotics/bootstrap}
\include{asymptotics/problems}

\input{common/postamble}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "asymptotics"
%%% End:
