\documentclass[12pt,notitlepage]{report}
\title{Core Econometrics}
\author{Gray Calhoun}
\date{10 Dec. 2013, version 0.7.1}

\input{tex/preamble}
\input{tex/macros}
\ifoptionfinal{
  % Set document to use fonts that may not be available locally
  \usepackage[bitstream-charter]{mathdesign}
  \usepackage[osf]{sourcecodepro} 
}{
  \tryinput{tex/localmod} % <- Put local modifications here
}

\begin{document}
\maketitle

\bigskip\noindent%
Copyright \textcopyright\ 2013--2014, the authors of \textit{Core Econometrics}.  A
complete author list is included in the Appendix titled ``Complete
list of authors.''

Permission is granted to copy, distribute and/or modify this document
under the terms of the GNU Free Documentation License, Version 1.3 or
any later version published by the Free Software Foundation; with no
Invariant Sections, no Front-Cover Texts, and no Back-Cover Texts.  A
copy of the license is included in the Appendix titled ``GNU Free
Documentation License'' and is also available online at
\url{http://www.gnu.org/copyleft/fdl.html}.

This text was produced as part of the Econometrics Free Library
Project.  The project's goal is to produce high-quality free and
open access econometrics textbooks and reference material.  More
information about this project is available at its homepage,
\url{http://www.econometricslibrary.org}, including information on how
to participate.  This text is typeset in LaTeX and there is a link
to the source of the document at the project's homepage as well.

Please cite this document as
\begin{quote}\raggedright
  G. Calhoun. \textit{Core Econometrics}. Econometrics Free Library
  Project, \url{http://www.econometricslibrary.org}, 2013, version
  0.7.1.
\end{quote}
For your convenience, we've provided the following BibTeX record.
\begin{Verbatim}
@Book{eflp-core,
author =    {Gray Calhoun},
title =     {Core Econometrics},
publisher = {Econometrics Free Library Project,
             \url{http://www.econometricslibrary.org}},
year =      {2013},
note =      {Version 0.7.1}}
\end{Verbatim}
When you find errors in this text, please email the project mailing
list,\footnote{\email{econometricslibrary@librelist.com}} or report
the error in the project's bug tracker.%
\footnote{\url{https://github.com/EconometricsLibrary/core/issues}}
It would be helpful if you could refer to the version number listed in
the date as well.  Thanks!

\tableofcontents
\addtocontents{toc}{\protect\setcounter{tocdepth}{1}}
\newcommand*\FancyVerbStopString{## end}
\newpage
\pagestyle{fancy}
\renewcommand{\chaptermark}[1]{\markboth{#1}{}} 
\lhead{\emph{Part \thepart, \parttitle}}
\rhead{\emph{\nouppercase \leftmark}}

\part{Probability overview}
\include{probability/random_variables}
\include{probability/properties}
\include{probability/bayesian}
\include{probability/asymptotics}
\chapter*{Additional exercises}
\addcontentsline{toc}{chapter}{Additional exercises}
\chaptermark{Exercises}
\input{probability/exercises}

\part{Estimation and inference}
\include{statistics/introduction}
\include{statistics/derivation}
\include{statistics/optimality}
\include{statistics/intervals}
\include{statistics/hypothesis_testing}
\include{statistics/likelihood_ratio_test}
\newcommand*\FancyVerbStartString{## block}
\include{statistics/bootstrap}
\include{statistics/multiple_testing}
\chapter*{Additional exercises}
\addcontentsline{toc}{chapter}{Additional exercises}
\chaptermark{Exercises}
\input{statistics/exercises}

\part{Statistical properties of regression}
\include{regression/introduction}
\include{regression/properties}
\include{regression/estimation}
\include{regression/inference}
\include{regression/asymptotics}
\chapter*{Additional exercises}
\addcontentsline{toc}{chapter}{Additional exercises}
\chaptermark{Exercises}
\input{regression/exercises}

\part{Interpreting and conducting a linear regression analysis}
\include{analysis/introduction}
\include{analysis/modeling}
\include{analysis/systems_of_equations}
\include{analysis/causality}
\chapter*{Additional exercises}
\addcontentsline{toc}{chapter}{Additional exercises}
\chaptermark{Exercises}
\input{analysis/exercises}

\addtocontents{toc}{\string\begin{NoHyper}}%
\addcontentsline{toc}{part}{Appendices}%
\addtocontents{toc}{\string\end{NoHyper}}
\lhead{}
% Don't put individual sections of the FDL in the table of contents
\addtocontents{toc}{\protect\setcounter{tocdepth}{0}}
\chapter*{Complete list of authors}
\addcontentsline{toc}{chapter}{Complete list of authors}
\input{AUTHORS}
\bibliography{tex/references}

\chapter*{GNU Free Documentation License}
\addcontentsline{toc}{chapter}{GNU Free Documentation License}
\chaptermark{GNU Free Documentation License}
\input{LICENSE}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "core_econometrics"
%%% End:
